---
title: "PIP Data Pipeline: Orchestration Overview"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article explains how the PIP data pipeline is orchestrated end to end:
which functions run in which order, which package owns each step, and how to
run a full session. Code in this article is illustrative and does not
execute when the article is built — running the pipeline requires a
configured working release, network access to Datalibweb, and write access
to PIP storage.

For the internal mechanics of DLW acquisition and validation, see the
companion article [Validating Data](Validating-Data.html); for survey
cleaning, deflation, and logging, see
[Processing Data functions](Processing-Data.html).

## Current pipeline entry points

The pipeline lifecycle uses three explicit entry points:

| Order | Wrapper | Package | Developer scope | Purpose |
|:-----:|---------|---------|------------------|---------|
| 1 | `update_aux_measures()` | **pipaux** | Auxiliary data engineer | Refresh auxiliary data (CPI, PPP, population, GDP, PCE, PFW) |
| 2 | `pipdata_dlw_process()` | **pipdata** | Survey ingestion engineer | Download new Datalibweb (DLW) survey files, validate them, update the DLW inventory |
| 3 | `pd_run_pipeline()` | **pipdata** | Survey processing engineer | Incrementally run clean, metadata, and deflate from completed validation state |

`update_aux_measures()` lives in the {pipaux} package and is not detailed in
this {pipdata} article beyond this table — it is a prerequisite step that
should run before Step 2 so that the latest auxiliary data is available.

`pipdata_dlw_process()` is the supported DLW acquisition/validation entry point.
`pd_run_pipeline()` is the top-level incremental entry point for the existing
durable processing stages. The older `pd_process_data()` and
`pd_deflate_pipeline()` wrappers remain independently callable and keep their
master-inventory returns.

**Important**: `pd_process_data()` cleans and attaches metadata to survey
data — it does **not** deflate welfare. Deflation is a separate, currently
manual step; see [Deflation is a separate step] below.

## How to run: an end-to-end session

### 0. Configure the working release

Start by checking for package updates, then configure the working release
that all three wrappers, and the functions that consume their output, read
from:

```{r setup}
# Check for updates to the PIP package ecosystem before running the pipeline
metapip::update_pip_packages()

release  <- "20260401"
identity <- "TEST"

pipfun::setup_working_release(release, identity, verbose = FALSE)
```

### 1. Refresh auxiliary data (pipaux)

```{r step1-aux}
# Owned by pipaux; see the pipaux package documentation for details.
pipaux::update_aux_measures()
```

### 2. Acquire and validate DLW data (pipdata)

```{r step2-dlw}
dlw_result <- pipdata::pipdata_dlw_process(
  inv_gmd_list      = "dlw_gmd_inv",
  get_dlw_data      = TRUE,
  validate_dlw_data = TRUE,
  check_missing     = TRUE,
  release           = release,
  identity          = identity
)
```

The wrapper requires explicit nonempty `release` and `identity` values. Their
formal `NULL` defaults are missing-required-value sentinels, not operational
defaults; `identity` is restricted to `"PROD"`, `"INT"`, or `"TEST"`.

This acquires five active modules and validates all seven recognized module
mappings, producing completed validation inventory (`gmd_valid_inv`) for Step
3. The assigned result is a plain aggregate. Inspect `dlw_result$outcome`
(`success`, `partial`, `failed`, or `no_work`) before deciding whether a script
should continue. Disabled or dependency-blocked nested stages use `not_run`.

Runtime stage failures after setup are represented in nested compact failure
tables and can yield partial or failed results. Invalid arguments, release
setup/precondition errors, explicit cancellation, and interrupts still escape.
Under automation a missing acquisition inventory returns failure rather than
opening a menu; validate-only execution never prompts. See
[Validating Data](Validating-Data.html) for the exact result and persistence
contracts.

### 3. Run clean, metadata, and deflate incrementally (pipdata)

```{r step3-process}
pipeline_result <- pd_run_pipeline(verbose = FALSE)
```

`pd_run_pipeline()` loads the completed validation inventory internally (via
`pipload::load_gmd_valid_inv()`) when `inv` is not supplied. It accepts current
`valid` and `invalid` available rows under existing cleaning policy, filters
recognized legacy blank/unavailable retry rows before planning, then runs only
stale or forced `clean:<survey_id>`, `metadata:<pip_id>`, and
`deflate:<pip_id>` nodes. It returns a compact `pipdata_pipeline_result`; it
does not return the master inventory. See
[Processing Data functions](Processing-Data.html) for the internal
mechanics.

Current nodes are cached. Stale or forced nodes are runnable. Failed
prerequisites make descendants blocked/skipped. `force_surveys` adds selected
survey or PIP chains without suppressing independent staleness. Unknown legacy
provenance requires explicit `bootstrap = TRUE`; use `bootstrap_entities` for a
restrictive canary.

The C2 manifest and exact Stamp receipts are the currentness authority. A
Colombia 2018 CPI change runs only matching Colombia 2018 metadata and deflate
nodes, while another Colombia year and unrelated country/year remain cached.
There is no persisted run cursor. After failure, call `pd_run_pipeline()` again;
it replans from the last valid manifest and exact Stamp facts.

### 4. Build a consolidated log report

```{r step4-report}
log_report(
  path      = file.path("log_reports", "log_report.md"),
  overwrite = TRUE
)
```

`log_report()` loads `"pipdata_log"` internally and summarizes DLW acquisition,
DLW validation, survey cleaning, deflation, and structured failures. DLW stage
sections use each latest attempt and exact completion metadata, with confined
fallback for older logs. DLW discriminators are excluded from generic type and
country tables to avoid double counting.

## Architecture and compatibility

Each stage is intentionally isolated because a different engineer developed it
independently, with its own auxiliary data, error handling, and logging:

- **`update_aux_measures()`** (pipaux) manages the auxiliary-data lifecycle:
  dependency resolution, GitHub/Y-drive sync, and change detection.
- **`pipdata_dlw_process()`** (pipdata) is responsible only for getting raw
  survey data into a validated state. It can continue validation after an
  acquisition failure when a trustworthy durable inventory remains, routes a
  custom inventory ID through both stages, and records summary/checkpoint facts.
  It does not clean or transform welfare data.
- **`pd_process_data()`** (pipdata) consumes the validated inventory and the
  refreshed auxiliary data, and is responsible for the cleaning/metadata
  transformation. Internally, it iterates one survey at a time, so that 
  a per-survey failure is caught, logged, and skipped without aborting
   the rest of the run.

The top-level executor composes these existing stage cores. Only clean,
metadata, and deflate are durable cached nodes. Load, PFW merge, recode,
auxiliary attachment, and save remain internal fingerprint components.

Each survey's cleaning pass inside `pd_process_data()` follows:
`inv_dlw_load()` (load raw survey) -> `pd_cpfw_merge()` (merge Price
Framework metadata) -> `pd_dlw_clean()` (S3-dispatched cleaning) ->
`pd_aux_attr()` (attach CPI/PPP/population/GDP/PCE metadata) ->
`save_pip_data()` (write cleaned data and metadata to PIP storage). No
deflation step is part of this chain.

## Deflation is a separate step

`pd_deflation()`/`deflation()` deflates a single cleaned survey's welfare
values using CPI, PPP, and population auxiliary data. It is **not** called
automatically by `pd_process_data()` — it must be run afterwards, per
survey, once cleaned data exists in PIP storage:

```{r deflation-example}
# Mode A: pass a cleaned survey directly (aux auto-loaded from the master
# inventory when cpi/ppp/pop are NULL)
dt <- pipload::pip_read(id = "BOL_2022_EH_INC_ALL", alias = "pip")
bol_deflated <- pd_deflation(dt)

# Mode B: load the survey by id and deflate in one call
bol_deflated <- pd_deflation(pip_id = "BOL_2022_EH_INC_ALL")
```

{pipload} also provides `load_pip_deflated_data()`, a convenience wrapper
around `pd_deflation()`'s Mode B: it locates a survey (by `id_name` or by
filter arguments such as `country_code`/`surveyid_year`/`module`), loads it,
and deflates it in one call — useful when you don't already have the
survey's `pip_id` in hand:

```{r deflation-pipload-example}
bol_deflated <- pipload::load_pip_deflated_data(id_name = "BOL_2022_EH_INC_ALL")

# Or filter by country/year/module instead of a known id_name
bol_deflated <- pipload::load_pip_deflated_data(
  country_code  = "BOL",
  surveyid_year = 2022,
  module        = "ALL"
)
```

See [Processing Data functions](Processing-Data.html) for details on both
modes.

## Production activation block

Local tests do not activate production. Production remains blocked until the
signed target Windows/SMB fencing and immutable unique-rename evidence are
complete.

## Logging and reporting scope

`log_report()` summarizes the shared `"pipdata_log"` entries from
`pipdata_dlw_process()` and `pd_process_data()`. It adds stage-aware warnings
for DLW-only, pipeline-only, incomplete, and DLW no-op runs. Acquisition and
validation are segmented independently at their latest attempt boundaries;
dedicated DLW sections own those records. Auxiliary refresh logging from
`pipaux` remains a separate follow-on scope.
