---
title: "PIP Data Pipeline: Orchestration Overview"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article explains how the PIP data pipeline is orchestrated end to end:
which functions run in which order, which package owns each step, and how to
run a full session. Code in this article is illustrative and does not
execute when the article is built — running the pipeline requires a
configured working release, network access to Datalibweb, and write access
to PIP storage.

For the internal mechanics of DLW acquisition and validation, see the
companion article [Validating Data](Validating-Data.html); for survey
cleaning, deflation, and logging, see
[Processing Data functions](Processing-Data.html).

## The three wrapper functions

The pipeline runs as three sequential stages, each owned by a dedicated
wrapper function in a separate package and (typically) a different
developer:

| Order | Wrapper | Package | Developer scope | Purpose |
|:-----:|---------|---------|------------------|---------|
| 1 | `update_aux_measures()` | **pipaux** | Auxiliary data engineer | Refresh auxiliary data (CPI, PPP, population, GDP, PCE, PFW) |
| 2 | `pipdata_dlw_process()` | **pipdata** | Survey ingestion engineer | Download new Datalibweb (DLW) survey files, validate them, update the DLW inventory |
| 3 | `pd_process_data()` | **pipdata** | Survey cleaning engineer | Merge auxiliary data with each DLW survey, clean variables, attach metadata, save versioned outputs, update the PIP master inventory |

`update_aux_measures()` lives in the {pipaux} package and is not detailed in
this {pipdata} article beyond this table — it is a prerequisite step that
should run before Step 2 so that the latest auxiliary data is available.

**Important**: `pd_process_data()` cleans and attaches metadata to survey
data — it does **not** deflate welfare. Deflation is a separate, currently
manual step; see [Deflation is a separate step] below.

## How to run: an end-to-end session

### 0. Configure the working release

Start by checking for package updates, then configure the working release
that all three wrappers, and the functions that consume their output, read
from:

```{r setup}
# Check for updates to the PIP package ecosystem before running the pipeline
metapip::update_pip_packages()

release  <- "20260401"
identity <- "TEST"

pipfun::setup_working_release(release, identity, verbose = FALSE)
```

### 1. Refresh auxiliary data (pipaux)

```{r step1-aux}
# Owned by pipaux; see the pipaux package documentation for details.
pipaux::update_aux_measures()
```

### 2. Acquire and validate DLW data (pipdata)

```{r step2-dlw}
pipdata::pipdata_dlw_process(
  inv_gmd_list      = "dlw_gmd_inv",
  get_dlw_data      = TRUE,
  validate_dlw_data = TRUE,
  log               = TRUE,
  save_log          = TRUE,
  check_missing     = TRUE,
  release           = release,
  identity          = identity
)
```

This downloads new/updated GMD survey files via `dlw::dlw_get_gmd()` and
validates them, producing an updated validated inventory
(`gmd_valid_inv`) consumed by Step 3. See
[Validating Data](Validating-Data.html) for the internal mechanics.

### 3. Clean surveys and build metadata (pipdata)

```{r step3-process}
new_pip_inv <- pd_process_data()
```

`pd_process_data()` loads the validated inventory internally (via
`pipload::load_gmd_valid_inv()`) when `inv` is not supplied, then iterates
it, merges auxiliary measures, cleans each survey, attaches metadata, saves
versioned outputs, and returns the updated PIP master inventory. See
[Processing Data functions](Processing-Data.html) for the internal
mechanics.

### 4. Build a consolidated log report

```{r step4-report}
log_report(
  path      = file.path("log_reports", "log_report.md"),
  overwrite = TRUE
)
```

`log_report()` loads the `"pipdata_log"` internally (via
`pipfun::log_filter(name = "pipdata_log")`) when `log` is not supplied, and
summarizes what `pd_process_data()` wrote to it — see
[Logging and reporting scope] below for what it does and does not cover.

## Architecture: why three wrappers?

Each stage is intentionally isolated because a different engineer developed it
independently, with its own auxiliary data, error handling, and logging:

- **`update_aux_measures()`** (pipaux) manages the auxiliary-data lifecycle:
  dependency resolution, GitHub/Y-drive sync, and change detection.
- **`pipdata_dlw_process()`** (pipdata) is responsible only for getting raw
  survey data into a validated state — it does not clean or transform
  welfare data.
- **`pd_process_data()`** (pipdata) consumes the validated inventory and the
  refreshed auxiliary data, and is responsible for the cleaning/metadata
  transformation. Internally, it iterates one survey at a time, so that 
  a per-survey failure is caught, logged, and skipped without aborting
   the rest of the run.

Each survey's cleaning pass inside `pd_process_data()` follows:
`inv_dlw_load()` (load raw survey) -> `pd_cpfw_merge()` (merge Price
Framework metadata) -> `pd_dlw_clean()` (S3-dispatched cleaning) ->
`pd_aux_attr()` (attach CPI/PPP/population/GDP/PCE metadata) ->
`save_pip_data()` (write cleaned data and metadata to PIP storage). No
deflation step is part of this chain.

## Deflation is a separate step

`pd_deflation()`/`deflation()` deflates a single cleaned survey's welfare
values using CPI, PPP, and population auxiliary data. It is **not** called
automatically by `pd_process_data()` — it must be run afterwards, per
survey, once cleaned data exists in PIP storage:

```{r deflation-example}
# Mode A: pass a cleaned survey directly (aux auto-loaded from the master
# inventory when cpi/ppp/pop are NULL)
dt <- pipload::pip_read(id = "BOL_2022_EH_INC_ALL", alias = "pip")
bol_deflated <- pd_deflation(dt)

# Mode B: load the survey by id and deflate in one call
bol_deflated <- pd_deflation(pip_id = "BOL_2022_EH_INC_ALL")
```

{pipload} also provides `load_pip_deflated_data()`, a convenience wrapper
around `pd_deflation()`'s Mode B: it locates a survey (by `id_name` or by
filter arguments such as `country_code`/`surveyid_year`/`module`), loads it,
and deflates it in one call — useful when you don't already have the
survey's `pip_id` in hand:

```{r deflation-pipload-example}
bol_deflated <- pipload::load_pip_deflated_data(id_name = "BOL_2022_EH_INC_ALL")

# Or filter by country/year/module instead of a known id_name
bol_deflated <- pipload::load_pip_deflated_data(
  country_code  = "BOL",
  surveyid_year = 2022,
  module        = "ALL"
)
```

See [Processing Data functions](Processing-Data.html) for details on both
modes.

## Logging and reporting scope

`log_report()` only parses the `"pipdata_log"` entries written by
`pd_process_data()`/`process_data()` (processing summary, per-survey
failures). It does **not** consume any log produced by
`pipdata_dlw_process()` (DLW acquisition/validation uses its own `log`/
`save_log` arguments and does not currently feed into `log_report()`). Keep
this scope in mind when reading a generated report — a clean report does not
imply the DLW acquisition/validation step had no issues.

The goal going forward is for `log_report()` to summarize a single,
harmonized log covering all three wrappers (`update_aux_measures()`,
`pipdata_dlw_process()`, and `pd_process_data()`), so one report reflects
the health of the entire pipeline run rather than just the cleaning step.
This is tracked on the roadmap (see the `unified-logging-report` idea).
