---
title: "Processing Data functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article is a deep dive into the third pipeline wrapper,
`pd_process_data()`, and the functions that consume its output:
`pd_deflate_pipeline()` (the second, batch-deflation stage),
`pd_deflation()`/`deflation()`, and `log_report()`. Code in this article is
illustrative and does not execute when the article is built — running it
requires a configured working release and write access to PIP storage.

For the end-to-end orchestration and how this wrapper fits with the other
two, see [PIP Data Pipeline: Orchestration Overview](PIP-data-pipeline.html).
For the internal mechanics of DLW acquisition and validation (the step
before this one), see [Validating Data](Validating-Data.html).

**Important**: `pd_process_data()` cleans surveys and attaches metadata —
it does **not** deflate welfare. Deflation is a separate pipeline stage:
use `pd_deflate_pipeline()` (below) to batch-deflate the whole inventory, or
`pd_deflation()` (see [Deflation is a separate step] below) for a single
survey.

## `pd_process_data()`: clean surveys and build metadata

`pd_process_data()` iterates a validated DLW inventory, merges auxiliary
data with each survey, cleans it, attaches metadata, saves versioned
outputs, and returns the updated PIP master inventory:

```{r pd-process-data}
new_pip_inv <- pd_process_data(
  inv          = NULL, # NULL loads pipload::load_gmd_valid_inv() internally
  aux_measures = c("pfw", "cpi", "ppp", "pop", "gdp", "pce"),
  force        = FALSE,
  verbose      = TRUE
)
```

- `inv` — the validated DLW inventory (as produced by
  [Validating Data](Validating-Data.html)'s `pipdata_validate_gmd()`). When
  `NULL` (the default), it is loaded internally via
  `pipload::load_gmd_valid_inv()`.
- `aux_measures` — which auxiliary datasets to load and merge/attach.
- `force` — when `TRUE`, temporarily switches `stamp` versioning to
  `"timestamp"` and bypasses the master-inventory comparison, forcing every
  survey in `inv` to be reprocessed.

### What happens inside

1. **Load auxiliary data** for each measure in `aux_measures` via
   `pipload::load_aux_data()`.
2. **Determine which surveys need (re-)processing** — `valid_dlw_load()`
   detects auxiliary-file changes (logging an `aux_changes_inf` entry when
   found), filters `inv` to the relevant modules
   (`ALL`/`GROUP`/`HIST`/`GPWG`/`BIN`), keeps the latest version of each
   survey, and — unless `force = TRUE` — drops surveys already present in
   the master inventory.
3. **Sync the recode spec once** via `sync_recode_spec()`, before the loop.
   This compares the package's bundled recode spec (see [What is the recode
   spec?] below) against the latest version saved in `stamp`: if there is no
   stamp version yet, or the package version has changed (compared by hash),
   it saves a new stamp version; otherwise it reuses the existing one.
   Either way it returns the active spec once, which is then threaded down
   to every survey's `pd_dlw_clean()` call — so the comparison/save only
   happens a single time per `pd_process_data()` run, instead of once per
   survey.
4. **Process each survey** with `process_data()`, one at a time:
   `inv_dlw_load()` (load raw survey) -> `pd_cpfw_merge()` (merge Price
   Framework metadata) -> `pd_dlw_clean()` (S3-dispatched cleaning,
   using the shared recode spec) -> `pd_aux_attr()` (attach
   CPI/PPP/population/GDP/PCE metadata) -> `save_pip_data(alias = "pip")` /
   `save_pip_data(alias = "pip_meta")` (write cleaned data and metadata to
   PIP storage). A per-survey failure is caught (as a `piperr` or generic
   error), logged to `"pipdata_log"` with the survey id and error class, and
   the survey is skipped — it does not abort the rest of the run. No
   deflation step is part of this chain.
5. **Log a processing summary** (`process_summary_inf`: totals,
   successes, failures) and, if any surveys failed, a `null_svys_inf` entry
   listing them.
6. **Rebuild the PIP master inventory** via `build_pip_inventory()` from the
   successful surveys' `stamp` catalog entries, and return it.

### What is the recode spec?

The **recode spec** is the set of recoding rules for the raw DLW variables
that `pd_dlw_clean()` cleans into PIP variables. It's defined in
`inst/extdata/recode_spec.yml`, bundled with the package: for each variable
it declares a `type`, a `recode_type` (one of `range_clamp`, `binary_map`,
`haven_labels`, `binned_from_continuous`, `quantile_from_continuous`, or
`indicator`), and the type-specific parameters that rule needs (e.g. a
`valid_range` for `range_clamp`, a `mapping` for `binary_map`/`haven_labels`).
`apply_recode_spec()` (called internally by `pd_dlw_clean()`) applies every
rule that matches a column present in the survey. Replace-type rules
(`range_clamp`, `binary_map`, `haven_labels`) recode the source column in
place (renaming it to the target variable name when they differ, e.g.
`urban` → `area`); derive-type rules (`binned_from_continuous`,
`quantile_from_continuous`) keep the source column and add the derived
variable alongside it (e.g. `age` stays, `age_group` is added).

## Deflation is a separate step

`pd_deflation()`/`deflation()` deflates a single cleaned survey's welfare
values using CPI, PPP, and population auxiliary data. It is **not** called
automatically by `pd_process_data()` — it must be run afterwards, per
survey, once cleaned data exists in PIP storage. There are two input modes:

```{r pd-deflation-modes}
# Mode A: pass a cleaned survey directly (aux auto-loaded from the master
# inventory when cpi/ppp/pop are NULL)
dt <- pipload::pip_read(id = "BOL_2022_EH_INC_ALL", alias = "pip")
bol_deflated <- pd_deflation(dt)

# Legacy Mode A: pass explicit aux tables instead of auto-loading them
ppp <- pipload::pip_load_aux("ppp")
cpi <- pipload::pip_load_aux("cpi")
pop <- pipload::pip_load_aux("pop")
bol_deflated <- pd_deflation(dt, cpi = cpi, ppp = ppp, pop = pop)

# Mode B: load the survey by id and deflate in one call
bol_deflated <- pd_deflation(pip_id = "BOL_2022_EH_INC_ALL")
```

In Mode A (`dt` supplied, `cpi`/`ppp`/`pop` left `NULL`), `pd_deflation()`
resolves the matching metadata version for the survey from the master
inventory and loads CPI/PPP/population automatically. In Mode B (`pip_id`
supplied instead of `dt`), the survey itself is also loaded from `stamp`
before deflation. `pd_deflation()` resolves the correct S3 method
(`deflation.pipmd()` for micro data, `deflation.pipgd()` for group data)
based on the survey's class and applies it; deflation failures are caught
and logged, returning `NA` for that survey rather than aborting.

{pipload} also provides `load_pip_deflated_data()`, a convenience wrapper
around `pd_deflation()`'s Mode B: it locates a survey (by `id_name` or by
filter arguments such as `country_code`/`surveyid_year`/`module`), loads it,
and deflates it in one call — useful when you don't already have the
survey's `pip_id` in hand:

```{r pipload-deflation-convenience}
bol_deflated <- pipload::load_pip_deflated_data(id_name = "BOL_2022_EH_INC_ALL")

# Or filter by country/year/module instead of a known id_name
bol_deflated <- pipload::load_pip_deflated_data(
  country_code  = "BOL",
  surveyid_year = 2022,
  module        = "ALL"
)
```

To deflate many surveys in a batch, use `pd_deflate_pipeline()`, the second
pipeline stage (see below), which loops `pd_deflation()` over the master
inventory.

## `pd_deflate_pipeline()`: batch-deflate the whole inventory

`pd_deflate_pipeline()` is the batch orchestrator that iterates over the PIP
master inventory of cleaned surveys, deflates each via `pd_deflation()`
(simple Mode B, `pip_id` only), detects and skips failures (including `NA`
returns), saves every deflated survey to the dedicated `"pip_deflated"` stamp
alias, updates the master inventory with deflation columns, and logs a
`deflate_summary_inf` summary entry:

```{r pd-deflate-pipeline}
new_pip_inv <- pd_deflate_pipeline(
  inv     = NULL, # NULL loads the master inventory internally
  force   = FALSE,
  verbose = TRUE
)
```

- `inv` — the master inventory (as produced by `pd_process_data()` →
  `build_pip_inventory()`). When `NULL` (the default), it is loaded internally
  via `pipload::load_pip_master_inventory()`.
- `force` — when `TRUE`, every row is re-deflated regardless of the `deflated`
  column. When `FALSE` (the default), only surveys whose `deflated` is `NA` or
  `FALSE` are deflated.
- `verbose` — controls verbosity of downstream `pipload` I/O calls.

**Relationship to `pd_process_data()`**: deflation is an independent second
stage. `pd_process_data()` cleans surveys and creates metadata; it does **not**
call `pd_deflate_pipeline()`. Run the pipeline wrapper explicitly after
cleaning (e.g. in `Pipdata_script.R`). The `"pip_deflated"` stamp alias must
be registered before the first run (e.g. via `pipdata_dlw_process()` or an
explicit `stamp::st_init()` pointing at `pip_repository/pip_deflated/`).

On success, the returned inventory has `deflated = TRUE`,
`content_hash_deflated` (the hash of the saved `"pip_deflated"` artifact), and
the `aux_cpi_hash_at_deflation`, `aux_ppp_hash_at_deflation`,
`aux_pop_hash_at_deflation` snapshots for each deflated survey. The
`"pipdata_log"` receives a `deflate_summary_inf` entry with `n_total`,
`n_success`, `n_failed`, `surveys_success`, and `surveys_failed`, which
`log_report()` renders as a *Deflation Summary* section.

## `log_report()`: summarize the `"pipdata_log"`

`log_report()` parses a `piplog` object and writes (or returns) a
structured markdown report:

```{r log-report}
# log defaults to pipfun::log_filter(name = "pipdata_log") when not supplied
report <- log_report()

# Or write directly to a file
log_report(
  path      = file.path("log_reports", "log_report.md"),
  overwrite = TRUE
)
```

The report includes, when the corresponding logmeta entry is present in the
log: DLW acquisition and validation summaries, stage-aware run warnings, run
metadata (time window, totals), the processing summary
(`process_summary_inf`), the deflation summary (`deflate_summary_inf`),
auxiliary file changes (`aux_changes_inf`), a summary table by error/info
type, a country-level breakdown of errors, inventory verification counts
(`inv_update_inf`), skipped-survey details
(`skipped_svys_data`/`skipped_svys_metadata`), and the list of surveys that
failed processing (`null_svys_inf`). Sections are silently omitted when
their logmeta entry is absent.

**Scope**: `log_report()` parses entries written under the `"pipdata_log"`
name by `pipdata_dlw_process()`, `pd_process_data()`/`process_data()`, and
`pd_deflate_pipeline()`/`deflate_one()`. The report is stage-aware and shows
DLW acquisition/validation outcomes together with cleaning and deflation
results. Auxiliary refresh logging from `pipaux` remains outside this report.

