---
title: "Validating Data: DLW Acquisition and Validation Internals"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article explains the internal mechanics of `pipdata_dlw_process()` —
the second of the three pipeline wrappers, responsible for downloading new
survey files from Datalibweb (DLW) and validating them before they are
handed off to survey cleaning. Code in this article is illustrative and
does not execute when the article is built — running it requires a
configured working release and network access to Datalibweb.

For the end-to-end orchestration and how this wrapper fits with the other
two, see [PIP Data Pipeline: Orchestration Overview](PIP-data-pipeline.html).
For what happens after validation (survey cleaning, deflation, logging), see
[Processing Data functions](Processing-Data.html).

## What `pipdata_dlw_process()` does

`pipdata_dlw_process()` is a thin orchestrator over two delegate functions,
run in order:

| Order | Function | Purpose |
|:-----:|----------|---------|
| 1 | `pipdata_get_gmd()` | Download new/updated GMD survey files from Datalibweb |
| 2 | `pipdata_validate_gmd()` | Validate the downloaded datasets and update the validated inventory |

Before either delegate runs, `pipdata_dlw_process()` ensures a working
release is configured and that a local GMD catalog list exists (downloading
it interactively via `dlw_gmd_list()` if missing).

```{r overview}
pipdata::pipdata_dlw_process(
  inv_gmd_list      = "dlw_gmd_inv",
  get_dlw_data      = TRUE,
  validate_dlw_data = TRUE,
  check_missing     = TRUE,
  release           = "20260401",
  identity          = "TEST"
)
```

Both `get_dlw_data` and `validate_dlw_data` can be set to `FALSE`
independently — for example, to re-run only validation without
re-downloading data.

## Step 1: Acquiring new datasets — `pipdata_get_gmd()`

`pipdata_get_gmd()` identifies datasets that are new or changed in the GMD
catalog and downloads them:

1. **Identify new datasets** — `dlw_gmd_new(check_missing, update_inventory = TRUE)`
   compares the local inventory against the GMD catalog and returns the rows
   that need downloading. If nothing is new, the function records a typed
   no-new-data entry and returns invisibly.
2. **Filter by module** — only `"ALL"`, `"GROUP"`, `"HIST"`, `"GPWG"`, and
   `"BIN"` modules are downloaded.
3. **Download per survey** — for each row, `dlw::dlw_get_gmd()` downloads the
   file (identified by country, year, survey acronym, module, and
   `vermast`/`veralt` version markers) to the local DLW data folder. Each
   download is wrapped in `tryCatch()`: failures are logged automatically with
   typed `logmeta` and the row is marked `data_available = "No"` rather than
   aborting the whole run; a success is marked `"Yes"`.

```{r step1-get-gmd}
pipdata::pipdata_get_gmd(
  inv_gmd_list  = "dlw_gmd_inv",
  check_missing = TRUE
)
```

## Step 2: Validating downloaded datasets — `pipdata_validate_gmd()`

`pipdata_validate_gmd()` runs after acquisition and validates the local
datasets that have not yet been validated:

1. **Identify unvalidated local datasets** — `dlw_gmd_unvalidated()` scans
   the local DLW data folder for datasets not yet recorded in the validated
   inventory.
2. **Load the existing validated inventory** — via
   `pipload::load_gmd_valid_inv()` (`gmd_valid_inv`), if one already exists.
3. **Diff against the existing inventory** — `gmd_to_validate()` identifies
   datasets that still need validation; `gmd_validated()` identifies datasets
   already validated in a prior run.
4. **Dispatch validation by module** — each dataset is validated with a
   module-specific function: `dlw_validation_gpwg()`, `dlw_validation_group()`,
   `dlw_validation_bin()`, `dlw_validation_hist()`, `dlw_validation_all()`,
   `dlw_validation_aspire()`, `dlw_validation_l()`, or `dlw_validation_skip()`
   as the default for unrecognized modules.
5. **Write the updated validated inventory** — the result becomes the new
   `gmd_valid_inv`, which `pd_process_data()` consumes as its starting
   inventory (see
   [Processing Data functions](Processing-Data.html)).

```{r step2-validate-gmd}
pipdata::pipdata_validate_gmd()
```

## Logging scope

Both delegates write unconditional typed entries to the shared
`"pipdata_log"` log. Acquisition uses `dlw_acquisition_inf`; validation uses
`dlw_validation_inf`. The enclosing `pipdata_dlw_process()` wrapper writes a
`dlw_summary_inf` stage marker and saves a DLW checkpoint automatically.
`log_report()` consumes these entries and presents DLW acquisition failures,
validation phases, and stage-aware warnings alongside the survey-cleaning
sections.
