---
title: "Validating Data: DLW Acquisition and Validation Internals"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article explains the internal mechanics of `pipdata_dlw_process()` —
the second of the three pipeline wrappers, responsible for downloading new
survey files from Datalibweb (DLW) and validating them before they are
handed off to survey cleaning. Code in this article is illustrative and
does not execute when the article is built — running it requires a
configured working release and network access to Datalibweb.

For the end-to-end orchestration and how this wrapper fits with the other
two, see [PIP Data Pipeline: Orchestration Overview](PIP-data-pipeline.html).
For what happens after validation (survey cleaning, deflation, logging), see
[Processing Data functions](Processing-Data.html).

## What `pipdata_dlw_process()` does

`pipdata_dlw_process()` is a thin orchestrator over two delegate functions,
run in order:

| Order | Function | Purpose |
|:-----:|----------|---------|
| 1 | `pipdata_get_gmd()` | Download new/updated GMD survey files from Datalibweb |
| 2 | `pipdata_validate_gmd()` | Validate the downloaded datasets and update the validated inventory |

Before either delegate runs, `pipdata_dlw_process()` ensures a working
release is configured and that a local GMD catalog list exists (downloading
it interactively via `dlw_gmd_list()` if missing).

```{r overview}
pipdata::pipdata_dlw_process(
  inv_gmd_list      = "dlw_gmd_inv",
  get_dlw_data      = TRUE,
  validate_dlw_data = TRUE,
  log               = TRUE,
  save_log          = TRUE,
  check_missing     = TRUE,
  release           = "20260401",
  identity          = "TEST"
)
```

Both `get_dlw_data` and `validate_dlw_data` can be set to `FALSE`
independently — for example, to re-run only validation without
re-downloading data.

## Step 1: Acquiring new datasets — `pipdata_get_gmd()`

`pipdata_get_gmd()` identifies datasets that are new or changed in the GMD
catalog and downloads them:

1. **Identify new datasets** — `dlw_gmd_new(check_missing, update_inventory = TRUE)`
   compares the local inventory against the GMD catalog and returns the rows
   that need downloading. If nothing is new, the function aborts with an
   informative message.
2. **Filter by module** — only `"ALL"`, `"GROUP"`, `"HIST"`, `"GPWG"`, and
   `"BIN"` modules are downloaded.
3. **Download per survey** — for each row, `dlw::dlw_get_gmd()` downloads the
   file (identified by country, year, survey acronym, module, and
   `vermast`/`veralt` version markers) to the local DLW data folder. Each
   download is wrapped in `tryCatch()`: a failure is logged (if `log = TRUE`)
   and the row is marked `data_available = "No"` rather than aborting the
   whole run; a success is marked `"Yes"`.

```{r step1-get-gmd}
pipdata::pipdata_get_gmd(
  inv_gmd_list  = "dlw_gmd_inv",
  log           = TRUE,
  save_log      = TRUE,
  check_missing = TRUE
)
```

## Step 2: Validating downloaded datasets — `pipdata_validate_gmd()`

`pipdata_validate_gmd()` runs after acquisition and validates the local
datasets that have not yet been validated:

1. **Identify unvalidated local datasets** — `dlw_gmd_unvalidated()` scans
   the local DLW data folder for datasets not yet recorded in the validated
   inventory.
2. **Load the existing validated inventory** — via
   `pipload::load_gmd_valid_inv()` (`gmd_valid_inv`), if one already exists.
3. **Diff against the existing inventory** — `gmd_to_validate()` identifies
   datasets that still need validation; `gmd_validated()` identifies datasets
   already validated in a prior run.
4. **Dispatch validation by module** — each dataset is validated with a
   module-specific function: `dlw_validation_gpwg()`, `dlw_validation_group()`,
   `dlw_validation_bin()`, `dlw_validation_hist()`, `dlw_validation_all()`,
   `dlw_validation_aspire()`, `dlw_validation_l()`, or `dlw_validation_skip()`
   as the default for unrecognized modules.
5. **Write the updated validated inventory** — the result becomes the new
   `gmd_valid_inv`, which `pd_process_data()` consumes as its starting
   inventory (see
   [Processing Data functions](Processing-Data.html)).

```{r step2-validate-gmd}
pipdata::pipdata_validate_gmd(
  log      = TRUE,
  save_log = TRUE
)
```

## Logging scope

Both delegates accept `log`/`save_log` arguments and write to the
`"pipdata_log"` log name via `pipfun::log_add()` — the same log name used by
`pd_process_data()`. However, this DLW acquisition/validation logging uses
an ad-hoc pattern (individual `log_add()` calls at each step) rather than
the structured `logmeta` summary entries (`process_summary_inf`,
`null_svys_inf`) that `pd_process_data()` emits. As a result,
`log_report()` — which parses those structured `logmeta` entries — does
**not** currently summarize DLW acquisition/validation activity, even though
both wrappers write to the same log name. Keep this in mind when interpreting
a generated report: it reflects survey cleaning only, not DLW acquisition or
validation outcomes.
